---
title: "Homework 7 - Predictive Modeling in Finance and Insurance"
author: "Dennis Goldenberg"
date: "2024-03-25"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = TRUE}
library(MASS)
library(ggplot2)
library(leaps)
Boston$chas <- factor(Boston$chas)
```

# 1. Generalized Linear Model

## a. Showing pdf belongs to exponential family

## b. Natural Exponential family or Exponential Dispersion Family

## c. Score statistic

## d. Pareto distribution and Exponential Family

\newpage

# 2. Logistic Regression

\newpage

# 3. Logistic Regression Part 2

## a. Coefficients from Models

## b. Probability Estimates from models

## c. Expeced Values from 3 Models

## d. Pearson residuals

## e. Goodness of fit statistics

\newpage

# 4. Maximum Likelihood Estimation approximation

## a. Plot of number of cases against time period

## b. Plot against log i

## c. Fitting GLM


\newpage

# 5. Deviance

## a. Calculating Deviance

## b. Calculating Pearson residual

## c. Calculating deviance residual

\newpage

# 6. Nominal Regression

## a. Odds ratio

## b. Probability Calculation

\newpage

# 7. Ordinal Regression
