# Homework 1 - Predictive Modeling in Finance and Insurance

# By: Dennis Goldenberg

```{r echo, include = TRUE}
# import packages
library(ggplot2)
library(patchwork)
```

## 1. Automobile Insurance Claims

```{r, include = TRUE}
# read in Data
auto <- read.table(file = 'AutoClaims-1.csv', header = TRUE, sep = ',')
auto$state <- factor(auto$state, ordered = TRUE)
auto$gender <- factor(auto$gender)
auto$class <- factor(auto$class)
```

### 1a.

I compute the descriptive statistics for the "PAID" variable:

```{r, inlcude = TRUE}
summary(auto$paid)
```

So, the mean paid is \$1,853.00 and the median paid is \$1,001.70.

### 1b.

I graph the histogram and qqplot of "paid", comparing quantiles:

```{r, include = TRUE}
hist <- ggplot(data = auto) + 
        geom_histogram(aes(paid), binwidth = 1000) +
        labs(title = "Histogram of Payouts") + 
        theme(plot.title = element_text(hjust = 0.5))
qq <- ggplot(data = auto) + geom_qq(aes(sample = paid)) +
      geom_qq_line(aes(sample = paid)) +
      labs(x = "Standard Normal Quantiles", y = 'Quantiles, Paid') + 
      labs(title = 'QQ plot') + 
      theme(plot.title = element_text(hjust = 0.5))
hist + plot_spacer() + qq + plot_layout(widths = c(6,0.5,6))
```

The qqplot suggests that the normal distribution is not a good fit. Based on the histogram, the distribution of paid looks $\textbf{lognormal}$.

### 1c.

I graph the histogram and qqplot of "logpaid", comparing quantiles:

```{r, include = TRUE}
hist <- ggplot(data = auto) + 
        geom_histogram(aes(logpaid), binwidth = 0.25) +
        labs(title = "Histogram of Payouts") + 
        theme(plot.title = element_text(hjust = 0.5))
qq <- ggplot(data = auto) + geom_qq(aes(sample = logpaid)) +
      geom_qq_line(aes(sample = logpaid)) +
    labs(x = "Standard Normal Quantiles", y = 'Quantiles, logPaid') +
      labs(title = 'QQ plot') + 
      theme(plot.title = element_text(hjust = 0.5))
hist + plot_spacer() + qq + plot_layout(widths = c(6,0.5,6))
```

Based on the good fit in the qqplot, and the general shape in the histogram, the distribution of "logpaid" seems to be $\textbf{approximately normal}$. This would suggest that the distribution of paid is $\textbf{lognormal}$.
