# Homework 1 - Predictive Modeling in Finance and Insurance

# By: Dennis Goldenberg

```{r echo, include = TRUE}
# import packages
library(GGally)
library(patchwork)
library(MASS)
```

## 1. Automobile Insurance Claims

```{r, include = TRUE}
# read in Data
auto <- read.table(file = 'AutoClaims-1.csv', header = TRUE, sep = ',')
auto$state <- factor(auto$state, ordered = TRUE)
auto$gender <- factor(auto$gender)
auto$class <- factor(auto$class)
```

### 1a.

I compute the descriptive statistics for the "PAID" variable:

```{r, inlcude = TRUE}
summary(auto$paid)
```

So, the mean paid is \$1,853.00 and the median paid is \$1,001.70.

### 1b.

I graph the histogram and qqplot of "paid", comparing quantiles:

```{r, include = TRUE}
hist <- ggplot(data = auto) + 
        geom_histogram(aes(paid), binwidth = 1000) +
        labs(title = "Histogram of Payouts") + 
        theme(plot.title = element_text(hjust = 0.5))
qq <- ggplot(data = auto) + geom_qq(aes(sample = paid)) +
      geom_qq_line(aes(sample = paid)) +
      labs(x = "Standard Normal Quantiles", y = 'Quantiles, Paid') + 
      labs(title = 'QQ plot') + 
      theme(plot.title = element_text(hjust = 0.5))
hist + plot_spacer() + qq + plot_layout(widths = c(6,0.5,6))
```

The qqplot suggests that the normal distribution is not a good fit. Based on the histogram, the distribution of paid looks $\textbf{lognormal}$.

### 1c.

I graph the histogram and qqplot of "logpaid", comparing quantiles:

```{r, include = TRUE}
hist <- ggplot(data = auto) + 
        geom_histogram(aes(logpaid), binwidth = 0.25) +
        labs(title = "Histogram of Payouts") + 
        theme(plot.title = element_text(hjust = 0.5))
qq <- ggplot(data = auto) + geom_qq(aes(sample = logpaid)) +
      geom_qq_line(aes(sample = logpaid)) +
    labs(x = "Standard Normal Quantiles", y = 'Quantiles, logPaid') +
      labs(title = 'QQ plot') + 
      theme(plot.title = element_text(hjust = 0.5))
hist + plot_spacer() + qq + plot_layout(widths = c(6,0.5,6))
```

Based on the good fit in the qqplot, and the general shape in the histogram, the distribution of "logpaid" seems to be $\textbf{approximately normal}$. This would suggest that the distribution of paid is $\textbf{lognormal}$.

### 1d.

I graph the scatterplot of "logpaid" against "age":

```{r, inlcude = TRUE}
scat <- ggplot(data = auto) + geom_point(aes(x = age, y = logpaid)) +
        labs(title = "Natural log of Paid Amount vs. age") +
        theme(plot.title = element_text(hjust = 0.5))
scat
```

This plot suggests that the variance of log payments for individuals on the younger side (closer to 50) is higher than the variance of payments for older individuals; outside of this, however, there appears to be $\textbf{no discernable relationship}$ between the natural log of the paid amount for an automobile insurance claim and the age of an individual.

### 1e.

I account for gender in the next scatter plot:

```{r, inlcude = TRUE}
scat <- ggplot(data = auto) + 
        geom_point(aes(x = age, y = logpaid, color = gender)) +
        labs(title = "Natural log of Paid Amount vs. age") +
        theme(plot.title = element_text(hjust = 0.5))
scat
```

Based on the scatter plot above, there is $\textbf{no discernable relationship}$ between the natural log of the paid amount for an automobile insurance claim and the gender of an individual.

## 2.

### 2a.

I first show to pairwise scatter plots of all combinations of variables:

```{r, include = TRUE}
pairs(Boston, col = 'blue')
```

This output is too small to discern any patterns; I look at the correlation matrix to see which scatter plots may have a strong relationship. I find where the correlations are strongest by looking to see where the absolute value of the correlation is greater than 0.4:

```{r, include = TRUE}
cormatrix <- cor(Boston)
strong_cor = cormatrix > 0.4 | cormatrix < -0.4
strong_cor
```
